---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
### (1) - Load data
```{r}
activity <- read.csv(unzip("activity.zip"), head=TRUE,sep=",")
```
### (2) - Check the data frame and transform date
```{r}
str(activity)
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?
### (1) - Histogram
```{r figure1}
agdata <- aggregate(activity$steps, by=list(date=activity$date), FUN=sum)
hist(agdata$x, main="Histogram of Total Number of steps per day",
    xlab="Total Number of Steps per day",
    ylab="Frequency",
    border="blue", 
    col="green",
    xlim=c(5000,25000),
    las=1, 
    breaks=10
)
```

### (2) - Mean and Median number of steps per day
```{r}
library(plyr)
avgdata <- ddply(activity, .(date), summarise,
               "Mean" = mean(steps),
               "Median" = median(steps)
)
print(avgdata)
```

## What is the average daily activity pattern?
### (1) - Time series plot
```{r figure2}
actintv <- aggregate(activity$steps, by=list(interval=activity$interval), FUN=mean, na.rm=TRUE)
colnames(actintv) <- c('interval', 'avgsteps')
with(actintv, plot(interval, avgsteps, type="l", main="Average Steps per 5-minute interval"))
```

### (2) - 5-minute interval with max number of steps
```{r}
actintv[actintv$avgsteps == max(actintv$avgsteps),]$interval
```

## Imputing missing values

### (1) - Total number of rows containing missing values
```{r}
nrow(activity[is.na(activity),])
```
### checking which columns actually contain missing values (nrow > 0)
```{r}
nrow(activity[is.na(activity$steps),])
nrow(activity[is.na(activity$date),])
nrow(activity[is.na(activity$interval),])
```
### (2) - Strategy for Filling in Missing values
#### We'll use the average steps per interval accross all days
####  Average values per interval come from previously created data frame (actintv)
### (3) - data frame with filled in mising values
```{r}
activit2 <- activity
for (i in 1:nrow(activit2)) {
    if (is.na(activit2[i,]$steps)){
        savintv = activit2[i,]$interval
        activit2[i,]$steps = actintv[actintv$interval == savintv,]$avgsteps
    }   
}
```
### (4) Aggregate the filled in data frame and make a Histogram
```{r figure3}
agdat2 <- aggregate(activit2$steps, by=list(date=activit2$date), FUN=sum)

hist(agdat2$x, 
     main="Histogram of Total Number of steps per day \n with Filled in Missing Values",
     xlab="Total Number of Steps per day",
     ylab="Frequency",
     border="blue", 
     col="green",
     xlim=c(5000,25000),
     las=1, 
     breaks=10
)
```

#### - Mean and Median number of steps per day
```{r}
avgact2 <- ddply(activit2, .(date), summarise,
               "Mean" = mean(steps),
               "Median" = median(steps)
)
print(avgact2)
```
#### Now comparing summarized data of NA filled IN and Not
```{r}
all.equal(avgdata,avgact2)
colMeans(avgdata[, c(2,3)], na.rm=TRUE)
colMeans(avgact2[, c(2,3)])
```
#### Per the result above the means did not change but the median did  


## Are there differences in activity patterns between weekdays and weekends?
### (1) - New `factor` variable with 2 levels "weekday"" and "weekend"  
#### create a vector and convert to `factor` specifying `levels/labels`
#### First, create a vector of weekdays
```{r}
activit2$date <- as.Date(activit2$date)
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activit2$wDay <- factor((weekdays(activit2$date) %in% weekdays1), 
                   levels=c(FALSE, TRUE), labels=c('weekend', 'weekday') 
                 )
```
### (2) Panel plot
#### summarize by interval and weekday factor
```{r}
avgdat2 <- ddply(activit2, .(interval, wDay), summarise,
                 "AvgSteps" = mean(steps)
            )

```
#### Time series plot
```{r figure4}
library(lattice)
xyplot(AvgSteps ~ interval | wDay, data=avgdat2,
       main = "Average number of steps per 5-minute Interval",
       xlab = "5-minute Interval", 
       ylab = "Average number of steps",
       type = c("l"), layout = c(2, 1)
)
```
